Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-08-12 analysis/paper/paper.Rmd
finish discussion and conclusion editing

The most abundant artifacts in the upper component are locally manufactured ceramics, which are distributed throughout the temporal sequence, and across the study area. More than 550,000 sherds were revealed, and around 1,200 vessels could be completely or partially reconstructed (i.e. complete rim or base). There are two shapes of locally-manufactured vessels; a cooking pot and a steamer made of two cooking pots stacked together with a clay filter between. Those vessel shapes demonstrate suites of standard morphological components. Each has a globular body with a short neck and wide mouth (Figure \@ref(fig:KWL-pots)). The exterior surface below the neck is decorated with a variety of impressed geometric motifs. These vessels were potentially used for cooking indicated by evidence of the frequent presence of charred residues and carbon deposits on vessel interiors, and soot on vessel exteriors. Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007]. The vessels were believed to be made with pinching technique according to the hand-shaped traces on vessel interiors, such as finger impressions and seams. This kind of vessel has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain [@NMTH2005].

@@

However, intentional standardization could also contribute to shape standardization in our case. To ensure the standardization is caused by a small number of specialized potters, we investigate other factors that may contribute to shape standardization, including the function and decoration of ceramics. To determine the function of pots in relation to shape changes, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909]. Stylish attributes present a wide variation in geometric motifs from pot to pot across three phases. In general, the ceramics in the Chinese period have slightly fewer types of motifs according to the ratio of distinct motifs to the total number for each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020]. The evidence shows that we could not completely rule out the intentional standardization in our sample.

The spatial pattern of ceramics could provide important information about the intensity and location of manufacturing or consumption. The spatial analysis (Figure \@ref(fig:kernel-plot)) reveals that pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in shape standardization. Also, the pottery manufacture inclines to be household-produced since no workshops have been found [@Chen2007].

@@

Our results offer tentative support for the hypothesis that foreign presence at Kiwulan influenced the emergence of social inequality in the local Indigenous society. If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Blanton1996; @Feinman1995; @Feinman2000; @Feinman2000political; @Feinman2010]. However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper. We find that vessel shapes were more standardized during the Chinese period than the European period.

@@

This study demonstrated the first use of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method for assessing differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan. Much lower variation in ceramic shape was identified during the period of Chinese presence. Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context. More homogeneous vessel shapes and sizes during the Chinese period, without any changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to intentional behavior instead of the emergence of specialized groups. Despite a small sample size in the Chinese period, the random distribution of ceramics does not support clustered patterns of manufacturing location, such as workshops. Instead, the production of ceramic is likely to be produced in households. The homogeneous appearance may suggest expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, maybe heightened during periods of foreign contact in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic assemblages. This study also broadens the GMM field by focusing on ceramic technologies, which could motivate more ceramic studies and become a promising branch parallel to current applications to lithic topology and bone morphology.

>>>>

The most abundant artifacts in the upper component are locally manufactured ceramics, which are distributed throughout the temporal sequence, and across the study area. More than 550,000 sherds were revealed, and around 1,200 vessels could be completely or partially reconstructed (i.e. complete rim or base). There are two shapes of locally-manufactured vessels; a cooking pot and a steamer made of two cooking pots stacked together with a clay filter between. Those vessel shapes demonstrate suites of standard morphological components. Each has a globular body with a short neck and wide mouth (Figure \@ref(fig:KWL-pots)). The exterior surface below the neck is decorated with a variety of impressed geometric motifs. These vessels were potentially used for cooking indicated by evidence of the frequent presence of charred residues and carbon deposits on vessel interiors, and soot on vessel exteriors. Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007]. The vessels were believed to be made with pinching technique according to some hand-shaped traces on vessel interiors, such as finger impressions and seams. This kind of vessel has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain [@NMTH2005].

@@

However, intentional standardization due to considerations of function or style could also contribute to the shape standardization in our case. To ensure shape standardization can reflect craft specialization, we investigated the function and surface decoration of the vessels. For function, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to explore foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909]. For style, surface impressed decoration usually consists of multiple bands of geometric motifs that we define as one type of decoration. If two pots share the same set of motifs but different arrangements of single bands, they are considered as two types. In general, the ceramics in the Chinese period have slightly fewer variations in decoration according to the ratio of distinct types to the total number of pottery from each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020]. The current evidence is not sufficient for us to completely rule out the possibility of intentional standardization.

The spatial pattern of ceramics could provide information about potential production units and production areas [@Costin2001]. The spatial analysis (Figure \@ref(fig:kernel-plot)) reveals that pottery samples present a widespread distribution with high densities of pottery at some units during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups [@Costin1995]. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in shape standardization. Also, it is more likely that ceramics are household-produced based on the spatial pattern and no specific facilities of production have been found [@Chen2007].

@@

Our results offer tentative support for the hypothesis that foreign presence at Kiwulan influenced the shape of vessels in the local Indigenous society. If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Blanton1996; @Feinman1995; @Feinman2000; @Feinman2000political; @Feinman2010]. However, strong claims for an emergence of social inequality resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper. We find that vessel shapes were more standardized during the Chinese period than the European period.

@@

This study demonstrated the first use of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method for assessing differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan. Much lower variation in ceramic shape was identified during the period of Chinese presence. Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context. More homogeneous vessel shapes and sizes during the Chinese period, without any substantial changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to intentional behavior instead of the emergence of specialized groups. Despite a small sample size in the Chinese period, the random distribution of ceramics does not support clustered patterns of manufacturing location, such as workshops. Instead, the ceramic production is likely to be manufactured in households. The homogeneous appearance may suggest expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, maybe heightened during periods of foreign contact in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic assemblages. This study also broadens the GMM field by focusing on ceramic technologies, which could motivate more ceramic studies and become a promising branch parallel to current applications to lithic typology and bone morphology.
